Legacy Data Confounds Genomics Studies

### 2019-06-20 LatexManuscript/legacy.tex
some small edits

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{../Figures/Figure1.jpg}
\caption{
Suspicious mutations carried by individuals with low quality data have distinct mutational profiles, reproduce poorly across studies, and are distributed across the genome. 
\textbf{A} 
Genome wide association of the average quality of mapped bases $Q$ for the 104 Japanese individuals included in the 1000 Genomes Project. 
This GWAS identified $587\ \  p < 10^{-8}$ and $1034\ \ p < 10^{-6}$ SNPs that were associated to the average $Q$ of SNPs mapped for an individual.
The same analysis was performed independently for each of the populations in the 1000 Genomes Project.
\textbf{B} 
Joint frequency spectrum plot of the Japanese from the 1000 Genomes Project and a more recent Japanese dataset from Nagahama.
The size of blue dots are proportional to the number of variants that associate with $Q$ in the JPT. 
\textbf{C} 
Mutation spectrum of the 1034 variants that associated with $Q$ in the JPT($p < 10^{-6}$), compared to a random set of SNPs.
The majority of the variants with significant associations to $Q$ have the *AC${\rightarrow}$*CC mutational pattern. 
There is also a slight enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations. 
These three enrichments can be summarized as G**${\rightarrow}$GG*. 
Stars ( * ) indicate significant difference from the mutational spectrum of a set of control SNPs.
}
 \label{SFS}
\end{figure}

@@

To identify SNPs that are likely to reproduce poorly across cohorts without having access to a second cohort, we performed an association study in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
Traditionally, genome wide association studies use genotypes as the independent variable. 
Here we perform a \todo{genotype conditional association test (GCAT), where} genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable \citep{song2015testing}.
We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$. 
While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$). 
The variants that are associated to $Q$ have a significant enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations (Figure \ref{SFS}C).
These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.
\todo{Statistical significance of these enrichment is computed using a chi-squared test as described by Harris 2017 \citep{Harris2017a}. }

@@

$Q$-associated variants are distributed across the genome, with chromosome 1 showing an excess of such variants, and other chromosomes being relatively uniform (Figure \ref{CountPerChrom}).
At a 10kb scale, we also see rather uniform distribution with a small number of regions showing an enrichment for such variants (Figure \ref{VariantDensity}).
An outlying 10kb region in chromosome 17 (bases : 22,020,000 to  22,030,000) has 35 $Q$-associated variants.
Distribution of association statistics in this region is provided in Figure \ref{Top10}.
By contrast, variants that do not pass the 1kGP strict mask are more unevenly distributed across the genome(Figure \ref{VariantDensity}).
\todo{The mutational 3-mer substitution patterns of this list of variants is similar to the signature identified in the single population test of the 1kGP JPT in that there is an enrichment in *AC${\rightarrow}$*CC.
However, another enrichment profile is significantly enriched : TAT${\rightarrow}$TTT, TCT${\rightarrow}$TTT and ACA${\rightarrow}$AAA (reverse complement : TGT${\rightarrow}$TTT). These three enrichments can be summarized as T*T${\rightarrow}$TTT.}

>>>>

\begin{figure}
\scalebox{0.85}{\includegraphics[width=\hsize,keepaspectratio]{../Figures/Figure1.jpg}}
\caption{
Suspicious mutations carried by individuals with low quality data have distinct mutational profiles, reproduce poorly across studies, and are distributed across the genome. 
\textbf{A} 
Genome wide association of the average quality of mapped bases $Q$ for the 104 Japanese individuals included in the 1000 Genomes Project. 
This GWAS identified $587\ \  p < 10^{-8}$ and $1034\ \ p < 10^{-6}$ SNPs that were associated to the average $Q$ of SNPs mapped for an individual.
The same analysis was performed independently for each of the populations in the 1000 Genomes Project.
\textbf{B} 
Joint frequency spectrum plot of the Japanese from the 1000 Genomes Project and a more recent Japanese dataset from Nagahama.
The size of blue dots are proportional to the number of variants that associate with $Q$ in the JPT. 
\textbf{C} 
Mutation spectrum of the 1034 variants that associated with $Q$ in the JPT($p < 10^{-6}$), compared to a random set of SNPs.
The majority of the variants with significant associations to $Q$ have the *AC${\rightarrow}$*CC mutational pattern. 
There is also a slight enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations. 
These three enrichments can be summarized as G**${\rightarrow}$GG*. 
Stars ( * ) indicate significant difference from the mutational spectrum of a set of control SNPs.
}
 \label{SFS}
\end{figure}

@@

To identify SNPs that are likely to reproduce poorly across cohorts without having access to a second cohort, we performed an association study in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
Traditionally, genome wide association studies use genotypes as the independent variable. 
Here we perform a \todo{genotype conditional association test (GCAT), where} genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable \citep{song2015testing}.
We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$. 
While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$). 
The variants that are associated to $Q$ have a significant enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations (Figure \ref{SFS}C).
These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.
\todo{Statistical significance of these enrichment is computed using a chi-squared test following \cite{Harris2017a}. }

@@

$Q$-associated variants are distributed across the genome, with chromosome 1 showing an excess of such variants, and other chromosomes being relatively uniform (Figure \ref{CountPerChrom}).
At a 10kb scale, we also see rather uniform distribution with a small number of regions showing an enrichment for such variants (Figure \ref{VariantDensity}).
An outlying 10kb region in chromosome 17 (bases : 22,020,000 to  22,030,000) has 35 $Q$-associated variants.
Distribution of association statistics in this region is provided in Figure \ref{Top10}.
By contrast, variants that do not pass the 1kGP strict mask are more unevenly distributed across the genome (Figure \ref{VariantDensity}).
\todo{The mutational 3-mer substitution patterns of this list of variants is similar to the signature identified in the single population test of the 1kGP JPT in that there is an enrichment in *AC${\rightarrow}$*CC.
However, another enrichment profile is significantly enriched : TAT${\rightarrow}$TTT, TCT${\rightarrow}$TTT and ACA${\rightarrow}$AAA (reverse complement : TGT${\rightarrow}$TTT). These three enrichments can be summarized as T*T${\rightarrow}$TTT.}
